What is happening on Twitter? A framework for student research projects with tweets

### 2020-06-02 Rmd/jse-special-issue-2020.Rmd
updated bib with blei ref on variational inference.

Our two students conducted their research projects during the 2015-2016 academic year. 
We recommend a full academic year for projects of this magnitude, although a one-semester 
project is possible. Our students presented their findings at the statistics department’s 
undergraduate poster session near the end of the 2015-2016 academic year.

@@

We include here instructions for creating a tweet collection. First, we created a 
new account on Twitter. With these user credentials, we used the 
R package `rtweet` to query the API. Because we work with linux operating systems, we used the `crontab` software 
to repeatedly execute R code to submit API queries. Each 
query lasted five minutes. We timed the API queries so 
that there was no time lag between queries. We stored tweets resulting from API queries in their native JSON format.

The R package `rtweet` 
provides functions that parse tweet JSON to R data frames. We then 
conducted all further analyses in R.

@@

Twitter API use agreements forbid users from sharing complete API query results. 
However, Twitter permits users to share tweet identification numbers. A user can 
then query a Twitter API to obtain complete tweet data. In our experience, this 
process is incomplete; that is, many tweets submitted to the Twitter API return 
no data. Additionally, on repeated querying of the API, different sets of tweets 
return data. This complicates our goal of making all analyses computationally 
reproducible.

From our collection of tweets, we chose to analyze those sent on three consecutive 
days from May 24, 2020 to May 26, 2020. We wanted to see if we could use text analysis 
tools to detect a transient change in topics for Memorial Day (May 25, 2020).

To minimize the computing requirements, we 
limited our analysis to tweets sent during a five-minute period (12:00pm to 12:05pm Eastern time) every day. However, our methods are appropriate for much larger data sets.
We then submitted API queries to Twitter
to get the full content of tweets, including the tweet text.
In supplementary files, we provide the R code that we used to 
query the Twitter API to obtain full tweet content.

@@

Tweets are available from the Twitter API as Javascript Object Notation (JSON) objects. Every tweet consists 
of multiple key-value pairs. The number of fields per 
tweet depends on user settings, retweet status, and other factors [@tweet_json]. 
The 31 tweet key-value pairs belong to
`r length(unique(data_dictionary$Type))` distinct classes (Appendix 1). The 
classes are either vectors - numeric, logical, or character - or arrays assembled 
from the vector classes.

@@

Our analyses use three fields from each tweet: date ("created_at"), tweet identifier 
("id_str"), and tweet text ("text"). The "created_at" field is a character string containing 
the date and time of the tweet. Every tweet has a unique identifier, the "id_str" value. The 
"text" field contains the unicode representation of the message. For our topic modeling, a single day defines a single "corpus", and a single tweet 
corresponds to a single "document".

@@

Latent Dirichlet allocation models require that the corpus be organized as a 
document by term matrix. In a document by term matrix, each row corresponds to a single document (a single 
tweet), and each column is a single term (or word). Each cell contains a 
count (the number of 
occurrences of a term in the specified document). We created a document by 
term matrix with the R function `cast_dtm` from the `tidytext` package.

@@

Inference for latent Dirichlet allocation model is performed by either 
sampling from the posterior distribution or through variational methods. Researchers
have devised a variety of Gibbs sampling techniques for latent Dirichlet allocation models. 
Variational methods, while using approximations to the posterior distribution, offer the 
advantage of computational speed. We used variational methods below.

@@

We sought to validate our hypothesis that we could detect a social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, we chose to 
analyze tweets before, during, and after Memorial Day (May 25, 2020). We fitted latent Dirichlet allocation models for each of three
distinct five-minute periods. The first period began at noon Eastern time on May 24, 2020. Subsequent time periods started 
24 and 48 hours later. 
We defined each time period to be a single collection, or corpus, of tweets. We then fitted latent 
Dirichlet allocation models to each corpus.

@@

Assigning meaning to topics is an active research area [@chang2009reading]. Since
our interest is in the transient appearance of a new topic, we don't attempt to
assign meaning to topics in our models.

We see that topic 7 from May 25 has several words that suggest Memorial Day: 
memorial, remember, honor, country. A similar topic is not seen on May 24 or May 26.

Some topics persist, possibly with distinct word probabilities, across the three days. For example, we see that President Trump features prominently in all three models.

We also note, on May 26, topic 10, which reflects discussion of the Amy Cooper Central Park incident.

We found evidence that latent Dirichlet allocation summarizes elements of 
large collections of Twitter conversations.

@@

In future iterations, we will use a written rubric to grade student theses. We'll
share the rubric with our students at the start of the academic year.

>>>>

Our two statistics students conducted their research projects during the 2015-2016 academic year. 
We recommend a full academic year for projects of this magnitude, although a one-semester 
project is possible. Our students presented their findings at the statistics department’s 
undergraduate poster session near the end of the 2015-2016 academic year.

We present below reproducible R code for analyzing data from May 2020. While these
are not the same data that our students analyzed in 2015, the methods and code 
are very similar.

@@

We include here instructions for creating a tweet collection. First, we created a 
new account on Twitter. With these user credentials, we used the 
R package `rtweet` to query the API. Because we work with linux operating systems, we used the `crontab` software 
to repeatedly execute R code to submit API queries. Each 
query lasted five minutes and produced a text file of approximately 130 MB. We timed the API queries so 
that there was no time lag between queries. We stored tweets resulting from API queries in their native JSON format.

@@

From our collection of tweets, we chose to analyze those sent on three consecutive 
days from May 24, 2020 to May 26, 2020. We wanted to see if we could use text analysis 
tools to detect a transient change in topics for Memorial Day (May 25, 2020). To minimize the computing requirements, we 
limited our analysis to tweets sent during a five-minute period (12:00pm to 12:05pm Eastern time) every day. However, our methods are appropriate for much larger data sets.
We then submitted API queries to Twitter
to get the full content of tweets, including the tweet text.
In supplementary files, we provide the R code that we used to 
query the Twitter API to obtain full tweet content.

Twitter API use agreements forbid users from sharing complete API query results. 
However, Twitter permits users to share tweet identification numbers. A user can 
then query a Twitter API to obtain complete tweet data. In our experience, this 
process is incomplete; that is, many tweets submitted to the Twitter API return 
no data. Additionally, on repeated querying of the API, different sets of tweets 
return data. This complicates our goal of making all analyses computationally 
reproducible and motivates our decision to share the tweet IDs of those tweets that 
we actually analyzed.

@@

Tweets are available from the Twitter API as Javascript Object Notation (JSON) objects [@json]. 
Every tweet consists 
of multiple key-value pairs. The number of fields per 
tweet depends on user settings, retweet status, and other factors [@tweet_json]. 
The 31 tweet key-value pairs belong to
`r length(unique(data_dictionary$Type))` distinct classes (Appendix 1). The 
classes are either vectors - numeric, logical, or character - or arrays assembled 
from the vector classes.

@@

Our analyses use three fields from each tweet: date ("created_at"), tweet identifier 
("id_str"), and tweet text ("text"). The "created_at" field is a character string containing 
the date and time of the tweet. Every tweet has a unique identifier, the "id_str" value. The 
"text" field contains the unicode representation of the message. After creating a text file 
with tweet JSON, our next step involved reading and parsing tweets with the R 
packages `rtweet` and `tidytext`.

@@

Latent Dirichlet allocation model fitting requires that the corpus be organized as a 
document by term matrix. In a document by term matrix, each row corresponds to a single document (a single 
tweet), and each column is a single term (or word). Each cell contains a 
count (the number of 
occurrences of a term in the specified document). We created a document by 
term matrix with the R function `cast_dtm` from the `tidytext` package.

@@

Inference for latent Dirichlet allocation model is performed by either 
sampling from the posterior distribution or through variational methods. Researchers
have devised a variety of Gibbs sampling techniques for these models. 
Variational methods, while using approximations to the posterior distribution, offer the 
advantage of computational speed [@blei2017variational]. We used variational methods below.

@@

We sought to validate our hypothesis that we could detect a social 
media event by examining tweet topic content at distinct time periods. 
As a proof of principle of our event detection strategy, we chose to 
analyze tweets before, during, and after Memorial Day (May 25, 2020). We fitted latent Dirichlet allocation models for each of three
distinct five-minute periods. The first period began at noon Eastern time on May 24, 2020. Subsequent time periods started 
24 and 48 hours later. 
We defined each time period to be a single collection, or corpus, of tweets.

@@

Assigning meaning to topics is an active research area [@chang2009reading]. Since
our interest is in the transient appearance of a new topic, we don't attempt to
assign meaning to every topic in our models. We see that topic 7 from May 25 has several 
words that suggest Memorial Day: memorial, remember, honor, country. A similar topic is not 
seen on May 24 or May 26. Some topics persist, possibly with distinct word probabilities, 
across the three days. For example, we see that President Trump features prominently in all 
three models. We also note, on May 26, topic 10, which reflects discussion of the Amy Cooper Central Park incident.

@@

```{r, summative, echo = FALSE, message = FALSE}
summative <- readr::read_csv("../data/summative-assessments.csv")
knitr::kable(summative, caption = "Rubric for summative assessment of learning objectives.") %>%
  kableExtra::column_spec(column = 1:5, width = "10em") %>%
  kableExtra::kable_styling(font_size = 6)

```

In future iterations, we will use a written rubric to grade student theses (Table \@ref(tab:summative)). We'll
share the rubric with our students at the start of the academic year.
